The role of Similarity in native spoken word recognition

### 2024-10-16 manuscript/manuscript.qmd
Fix figures

```{r}
#| label: setup
library(targets)
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(gt)
library(gtExtras)
library(brms)
library(tidybayes)

@@

When some German speakers listened to the song *The Power* by SNAP! (*Coyote Ugly*, 2000), many of them misheard the line "I've got the power" as *Agatha Bauer!* When Michael Jackson vocalised "ma-ma-coo-sah" in his song *Wanna Be Startin' Somethin'* (*Thriller*, 1982), some Dutch speakers understood *mama appelsap* [mama applejuice]. When Spanish speakers listen to the line "Circumvent your thick ego" from the song *Pictures* by System of a Down (*Steal this Album!*, 2002), they tend to hear *Sácame de aquí, King-Kong* [take me out of here, King-Kong]. When Japanese speakers listen to the line "I want to hold your hand" in the homonym song by The Beatles (*Meet The Beatles*, 1964), they often mishear *Aho na hōnyōhan* [stupid urinator]. Outrageous as these examples might sound to speakers of other languages, many readers may know a few cases in their own native language. This auditory illusion can feel quite real, and often inevitable after the first time it happens. In Japanese, this phenomenon takes the name *Soramimi* (lit. "empty ear"), after the *Soramimi Hour*---a section of the program *Tamori Club* hosted by comedian Morita Kazuyoshi (Tamori)---in which instances of this illusion were presented with comedy purposes. Soramimi is a particular case of *homophonic translation*: words or phrases in one language are translated into similar-sounding words and phrases in a different language, without necessarily preserving the meaning. This phenomenon extends beyond song lyrics, and may occur anytime when listening to non-native speech. For instance, when listening to "You know me?" (\textipa{/ju: n@U mi:}/) in English, Japanese-native listeners tend to understand *yunomi* (/\textipa{jUnomi}/), which is a type of Japanese teacup. Overall, the occurrence of Soramimi---and more generally, homophonic translations---suggests that speech in an unfamiliar language has the potential to activate lexical representations in the native language.

@@

@@

After data collection, participants' answers were manually coded into the following categories: *Correct*, *Typo*, *Wrong*, *False friend*, *Other*. A response was coded as *Correct* if the provided string of characters was identical to the orthographic form of the correct translation. A response was coded as *Typo* if the participant provided a string of characters only one edit distance (addition, deletion, or substitution) apart from the orthographic form of the correct translation (e.g., "pengiun" instead of "penguin"), as long as the response did not correspond to a distinct English word. A response was coded as *False friend* if the participant provided a phonologically similar incorrect translation. Responses not meeting the criteria for previous categories were labelled as *Wrong* or *Other* (see Data analysis section for more details). Both *Correct* and *Typo* responses were considered as correct, while *Wrong* and *False friend* responses were considered as incorrect. *Other* responses were excluded from data analysis. Trials in which participants took longer than 10 seconds to respond were also excluded. Participants contributed a total of `r nrow(exp_responses)` valid trials (`r format(nrow(exp_responses[exp_responses$group %in% c("cat-ENG", "cat-SPA"),]), big.mark = ",")` in Catalan, `r format(nrow(exp_responses[exp_responses$group %in% c("spa-ENG"),]), big.mark = ",")` in Spanish). The task took approximately 15 minutes to be completed.

@@

::: {.content-visible when-format="pdf"}
$$
\begin{aligned}
&\textbf{Likelihood}  \\
y_{i} \sim & \text{Bernoulli}(p_{i}) \\ \\
&\textbf{Parameters}  \\
\text{Logit}(p_{i}) = &  \beta_{0[p,w]} + \beta_{1[p]} \text{Frequency}_{i} + \beta_{2[p]} \text{PTHN}_i + \beta_{3[p]} \text{Similarity}_i + \beta_{4[p]} (\text{PTHN}_i \times \text{Similarity}_i) \\
\beta_{0-6[p,w]} \sim & \mathcal{N}(\mu_{\beta_{j}}, \sigma_{\beta_{j}}) \text{, for participant } p \text{ in 1, ..., } P \text{ and  word } w \text{ in 1, ..., } W \\
\beta_{1-6[p]} \sim &  \mathcal{N}(\mu_{\beta_{j}}, \sigma_{\beta_{j}}) \text{, for participant } p \text{ in 1, ..., } P \\ \\
&\textbf{Prior}  \\
\mu_{\beta_{p,w}}  \sim &  \mathcal{N}(0, 0.1) \\
\sigma_{\beta_{p}},  \sigma_{\beta_{w}} \sim & \text{HalfCauchy}(0, 0.1) \\
\rho_{p}, \rho_{w} \sim & \text{LKJCorr}(8) \\
\end{aligned}
$$
::: {eq-1}

To test the practical relevance of each predictor we followed @kruschke2018bayesian. We first specified a region of practical equivalence (*ROPE*) around zero ([-0.1, +0.1], in the logit scale). This area indicates the values of the regression coefficients that we considered as equivalent to zero. We then computed the 95% posterior credible intervals (CrI) of the regression coefficients of interest. Finally, we calculated the proportion of the 95% CrI that fell inside the ROPE. This proportion indicates the probability that the true value of the coefficient is equivalent to zero. All analyses were performed in R environment [@rcoreteam2013language]. We used the tidyverse family of R packages [@wickham2019welcome] to process data and to generate figures. We used the `brms` R package [@burkner2017brms] using the `cmdstanr` back-end to the Stan probabilistic language [@carpenter2017stan] to estimate and compare the models (see Appendix 1 for mode details on the models).

@@

We collected data for a total of `r format(nrow(dataset$cat_eng) + nrow(dataset$spa_eng), big.mark = ",")` trials completed by `r n_participants$cat_eng + n_participants$spa_eng` unique participants. Of those trials, `r format(nrow(dataset$cat_eng), big.mark = ",")` were provided by `r n_participants$cat_eng` unique participants who listened to Catalan words, and `r format(nrow(dataset$spa_eng), big.mark = ",")` trials were provided by `r n_participants$spa_eng` unique participants who listened to Spanish words. We excluded trials in which participants did not enter any text (*n* = `r r_validity$cat_eng["blank"]`), in which a response in a language other than English was provided (e.g., `agua`, *n* = `r r_validity$cat_eng["language"]`), in which participants did not provide a whole word (e.g., `f`, *n* = `r r_validity$cat_eng["incomplete"]`), and in which participants added comments to the experimenter (e.g., `unsure`, *n* = `r r_validity$cat_eng["other"]`). In addition, we excluded data from participants that self-rated their oral and/or written skills in Catalan and Spanish, or any other second language as four or higher in a five-point scale (*n* = `r excluded_lang`), were diagnosed with a language (*n* = `r excluded_impairment`), or did not contribute more than 80% of valid trials (*n* = `r excluded_ntrials`).

@@

@tbl-dataset shows a summary of participants' accuracy across Experiments 1, 2, and 3. MCMC chains in the model showed strong evidence of convergence ($\hat{R}<1.01$) (see Appendix 2 for more detailed model diagnostics). Participants translating Catalan words and participants translating Spanish words performed equivalently, as indicated by the regression coefficient of *Group* ($\beta$ = `r round(c1$b_group1$.median, 3)`, 95% CrI = [`r round(c1$b_group1$.lower, 3)`, `r round(c1$b_group1$.upper, 3)`], *p*(ROPE) = `r round(c1$b_group1$.rope, 3)`). Overall, participants responded less accurately to words with more CLPNs than to words with fewer CLPNs, regardless of the amount of phonological similarity between the presented word and its translation. This is indicated by the size of the regression coefficient of the two-way interaction between *Similarity* and *CLPN* ($\beta$ = `r round(c1$b_neigh_n_h_lv$.median, 3)`, 95% CrI = [`r round(c1$b_neigh_n_h_lv$.lower, 3)`, `r round(c1$b_neigh_n_h_lv$.upper, 3)`], *p*(ROPE) = `r round(c1$b_neigh_n_h_lv$.rope, 3)`). As anticipated, participants' performance benefited from an increase in *Similarity* ($\beta$ = `r round(c1$b_lv$.median, 3)`, 95% CrI = [`r round(c1$b_lv$.lower, 3)`, `r round(c1$b_lv$.upper, 3)`], *p*(ROPE) = `r round(c1$b_lv$.rope, 3)`), while the number of *CLND* had the opposite effect ($\beta$ = `r round(c1$b_neigh_n_h$.median, 3)`, 95% CrI = [`r round(c1$b_neigh_n_h$.lower, 3)`, `r round(c1$b_neigh_n_h$.upper, 3)`], *p*(ROPE) = `r round(c1$b_neigh_n_h$.rope, 3)`). @fig-epreds-1 illustrates the posterior of the average predictions of the model for words with different values of *Similarity* and *CLPN*. @fig-coefs shows a graphic summary of the posterior distribution of the regression coefficients of interest.

@@

We collected data for a total of `r format(nrow(dataset$cat_spa) + nrow(dataset$cat_spa), big.mark = ",")` trials completed by `r n_participants$cat_spa` unique participants. We excluded trials in which participants did not enter any text (*n* = `r r_validity$cat_spa["blank"]`), in which a response in a language other than English was provided (e.g., `agua`, *n* = `r r_validity$cat_spa["language"]`), in which participants did not provide a whole word (e.g., `f`, *n* = `r r_validity$cat_spa["incomplete"]`), and in which participants added comments to the experimenter (e.g., `unsure`, (*n* = `r r_validity$cat_spa["other"]`). In addition, we excluded data from participants that self-rated their oral and/or written skills in Catalan and Spanish, or any other second language as four or higher in a five-point scale (*n* = `r excluded_lang`), were diagnosed with a developmental language disorder (*n* = `r excluded_impairment`), or did not contribute more than 80% of valid trials (*n* = `r excluded_ntrials`). After applying trial-level and participant-level inclusion criteria, the resulting dataset included `r format(nrow(dataset_clean$cat_spa), big.mark = ",")` trials provided by `r n_distinct(dataset_clean$cat_spa$participant_id) + n_distinct(dataset_clean$cat_spa$participant_id)` unique participants. Of those trials, `r format(nrow(dataset_clean$cat_spa), big.mark = ",")` were provided by `r n_distinct(dataset_clean$cat_spa$participant_id)` unique participants who listened to Catalan words, and `r format(nrow(dataset_clean$cat_spa), big.mark = ",")` trials were provided by `r n_distinct(dataset_clean$cat_spa$participant_id)` unique participants who listened to Spanish words. Responses given by participants were `r round(lengths$len_mean, 2)` characters long on average (*Median* = `r round(lengths$len_median, 2)`, *SD* = `r round(lengths$len_sd, 2)`, Range = `r round(lengths$len_min, 2)`-`r round(lengths$len_max, 2)`).

```{r}
c2 <- gather_draws(exp_2_m0, `b_.*`, `sd_.*`, regex = TRUE) |>
  summarise(across(.value, lst(
    .median = median,
    .lower = \(x) hdi(x)[1],
    .upper = \(x) hdi(x)[2],
    .rope = \(x) mean(between(x, -0.1, 0.1))
  ),
  .names = "{fn}"
  ))
c2 <- split(c2, c2$.variable)
names(c2) <- janitor::make_clean_names(names(c2))
```

@@

```{r tbl-surprises}
#| label: tbl-surprises
#| tbl-cap: "List of items with unexpectedly high accuracy: the Levenshtein similarityscore betwen the presented word (in Catalan or Spanish) and their correct Enlgish translation is zero, but participants, who are reportedly unfamiliar with the presented language, were on average >10% likely to guess the correct translation."
tbl_data <- list(
  "Experiment 1" = dataset_1,
  "Experiment 2" = dataset_2
) |>
  bind_rows(.id = "exp") |>
  summarise(
    n = n(),
    correct = sum(correct),
    .by = c(exp, group, lv, translation)
  ) |>
  mutate(
    accuracy = prop_adj(correct, n),
    accuracy_se = prop_adj_se(correct, n)
  ) |>
  dplyr::filter(lv == 0, accuracy >= 0.1) |>
  select(exp, translation, group, accuracy, accuracy_se) |>
  arrange(exp, group, desc(accuracy))

tbl_data |>
  gt(
    rowname_col = "translation",
    groupname_col = c("exp", "group")
  ) |>
  fmt_number(is.numeric, scale_by = 100) |>
  cols_label(accuracy = "Accuracy (%)", accuracy_se = "SE")
```

@@

[ADD HERE LEXICAL FREQUENCY OF CATALAN AND SPANISH WORDS IN ENGLISH, AND OF CATALAN WORDS IN SPANISH]

@@

```{r}
#| label: participants-excluded-3
excluded_lang <- participants_3 |>
  filter(l2_writ_prod > 3 |
    l2_oral_comp > 3 |
    cat_oral_comp > 3 |
    cat_writ_prod > 3 |
    spa_oral_comp > 3 |
    spa_writ_prod > 3) |>
  nrow()

@@

```{r dataset-3}
#| label: dataset-3
dataset <- quest_responses |>
  filter(group != "cat-SPA") |>
  group_split(group) |>
  set_names(c("spa-ENG", "cat-ENG")) |>
  janitor::clean_names()

dataset <- split(quest_responses, quest_responses$group)
names(dataset) <- janitor::make_clean_names(names(dataset))
r_validity <- map(
  dataset,
  \(x) janitor::clean_names(table(x$response_type))
)
n_participants <- map(dataset, \(x) n_distinct(x$participant_id))

@@

We collected data for a total of `r format(nrow(dataset$cat_eng) + nrow(dataset$spa_eng), big.mark = ",")` trials completed by `r n_participants$cat_eng + n_participants$spa_eng` unique participants. Of those trials, `r format(nrow(dataset$cat_eng), big.mark = ",")` were provided by `r n_participants$cat_eng` unique participants who listened to Catalan words, and `r format(nrow(dataset$spa_eng), big.mark = ",")` trials were provided by `r n_participants$spa_eng` unique participants who listened to Spanish words. We excluded trials in which participants did not enter any text (*n* = `r dataset$cat_eng$blank`), in which a response in a language other than English was provided (e.g., `agua`, *n* = `r dataset$cat_eng$language`), in which participants did not provide a whole word (e.g., `f`, *n* = `r dataset$cat_eng$incomplete`), and in which participants added comments to the experimenter (e.g., `unsure`, (*n* = `r dataset$cat_eng$other`). In addition, we excluded data from participants that self-rated their oral and/or written skills in Catalan and Spanish, or any other second language as four or higher in a five-point scale (*n* = `r excluded_lang`), were diagnosed with a language (*n* = `r excluded_impairment`), or did not contribute more than 80% of valid trials (*n* = `r excluded_ntrials`).

@@

```{r}
lengths <- dataset_3 |>
  filter(!knowledge) |> 
  mutate(len = map_int(strsplit(response, ""), length)) |>
  summarise(
    across(
      len,
      lst(mean, median, sd, min, max)
    ),
    .by = c(group)
  )
lengths <- split(lengths, lengths$group)

@@

```{r fig-knowledge}
#| label: fig-knowledge
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "Catalan/Spanish prior word knowledge as reported by English native participants in Experiment 3. (A) Average proportion of participants that reported prior knowledge for words with surprisingly high accuracy (no phonological similarity with the correct translation, accuracy higher than 10%), and for words with expected accuracy (low similarity-low accuracy, or high similarity-high accuracy). (B) Average accuracy for words with expected and unexpected accuracy."
plot_1_data <- dataset_3 |>
  mutate(surprise = translation %in% tbl_data$translation) |>
  summarise(
    knowledge = mean(knowledge),
    .by = c(translation, group, surprise)
  ) |>
  select(translation, group, surprise, knowledge) |>
  mutate(surprise = if_else(
    surprise,
    "Unexpectedly high accuracy\n(no phonological similarity)",
    "Expected accuracy"
  ))

plot_1 <- plot_1_data |> 
  filter(surprise=="Unexpectedly high accuracy\n(no phonological similarity)") |> 
  ggplot(aes(1, knowledge, colour = group, fill = group)) +
  facet_wrap(~group, ncol = 1) +
  geom_dots(aes(group = NA), side = "both", shape = 1, layout = "swarm", stackratio = 1.5, dotsize = 1, binwidth = 0.05, show.legend = FALSE) +
  geom_rect(
    data = filter(plot_1_data, surprise=="Expected accuracy") |> 
      summarise(
      knowledge_mean = mean(knowledge), 
      knowledge_std = sd(knowledge),
      n = n(),
      .by = group
    ) |> 
    mutate(knowledge_se = knowledge_std / sqrt(n)),
   aes(xmin = 0.5, xmax = 1.5, 
   ymin = knowledge_mean - knowledge_se, 
   ymax = knowledge_mean + knowledge_se, 
   fill = group, 
   colour = group
   ), 
       inherit.aes = FALSE,
   alpha = 1/4,
   color = NA) +
  geom_hline(
    data = filter(plot_1_data, surprise=="Expected accuracy") |>
      summarise(knowledge_mean = mean(knowledge), .by = group),
    aes(yintercept = knowledge_mean, color = group)
  ) +
  labs(
    y = "Prior knowledge reported",
    x = "Group",
    colour = "Group",
    fill = "Group"
  ) +
  scale_y_continuous(labels = scales::percent) +
  theme(axis.text.x = element_blank(),
  axis.title.x = element_blank())

>>>>

```{r}
#| label: setup
library(targets)
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(gt)
library(gtExtras)
library(brms)
library(tidybayes)
library(ggExtra)
library(tinytable)

@@

When some German speakers listened to the song *The Power* by SNAP! (*Coyote Ugly*, 2000), many of them misheard the line "I've got the power" as *Agatha Bauer!* When Michael Jackson vocalised "ma-ma-coo-sah" in his song *Wanna Be Startin' Somethin'* (*Thriller*, 1982), some Dutch speakers understood *mama appelsap* [mama applejuice]. When Spanish speakers listen to the line "Circumvent your thick ego" from the song *Pictures* by System of a Down (*Steal this Album!*, 2002), they tend to hear *Sácame de aquí, King-Kong* [take me out of here, King-Kong]. When Japanese speakers listen to the line "I want to hold your hand" in the homonym song by The Beatles (*Meet The Beatles*, 1964), they often mishear *Aho na hōnyōhan* [stupid urinator]. Outrageous as these examples might sound to speakers of other languages, many readers may know a few cases in their own native language. This auditory illusion can feel quite real, and often inevitable after the first time it happens. In Japanese, this phenomenon takes the name *Soramimi* (lit. "empty ear"), after the *Soramimi Hour*---a section of the program *Tamori Club* hosted by comedian Morita Kazuyoshi (Tamori)---in which instances of this illusion were presented with comedy purposes. Soramimi is a particular case of *homophonic translation*: words or phrases in one language are translated into similar-sounding words and phrases in a different language, without necessarily preserving the meaning [@gasparov2006semen]. This phenomenon extends beyond song lyrics, and may occur anytime when listening to non-native speech. For instance, when listening to "You know me?" (\textipa{/ju: n@U mi:}/) in English, Japanese-native listeners tend to understand *yunomi* (/\textipa{jUnomi}/), which is a type of Japanese teacup. Overall, the occurrence of Soramimi---and more generally, homophonic translations---suggests that speech in an unfamiliar language has the potential to activate lexical representations in the native language.

@@

We further investigated the effect of phonological competitors. Even if the presented word and its target translation share high phonological similarity, participants may still provide incorrect translations if the presented word also shares high phonological similarity with other words in the native language. Words with denser phonological neighbourhoods are recognised more slowly and less accurately than words with sparser phonological neighbourhoods [@dufour2003lexical; @goldinger1989priming; @hamburger1996phonological; @luce1990similarity; @luce1998recognizing]. This is especially true if such neighbours share higher phonological similarity with the presented word, or are lexically more frequent than the target translation [@luce1998recognizing; @dufour2010phonological]. Since the interference effects of phonological neighbourhood density has also been reported across languages [@weber2004lexical], we hypothesised that the size of the facilitatory role of phonological similarity would be inversely proportional to the amount of higher-frequency phonological neighbours of the presented word in the target language. In the case of non-native speech recognition, the presence of cross-linguistic pairs which are phonologically similar but differ in meaning (e.g., false friends) may act as distractors during lexical access, obstructing the selection of the appropriate target translation in the listener’s lexicon. For instance, @otwinowska2019more reported that false friends were disadvantaged relative to non-cognates by Polish second language learners, in contrast to cognates which were known better. It is therefore important to investigate the joint effect of both cognates and false friends when investigating the effect cross-linguistic phonological similarity has on word recognition in a foreign language. This could shed light on available strategies and challenges associated with the early stages of second language acquisition. For instance, one could expect English participants to incorrectly translate the Spanish word *botón* as *bottom* instead of as its correct translation *button*, due to combined effect of the close phonological similarity between *botón* and *bottom* along with the high lexical frequency of *bottom* relative to *button*. To test this prediction, we developed a lexical frequency-dependent measure of cross-language phonological neighbourhood density, in which a neighbour is counted only if it is higher frequency and is one phoneme apart from the presented word. If the phonological neighbourhood density of the target translation affects participants’ performance negatively, this would suggest that competitors in the native language affect recognition of non-native words during recognition of foreign speech. We conducted a series of three experiments to test these predictions.

@@

After data collection, participants' answers were manually coded into the following categories: *Correct*, *Typo*, *Wrong*, *False friend*, *Other*. A response was coded as *Correct* if the provided string of characters was identical to the orthographic form of the correct translation. A response was coded as *Typo* if the participant provided a string of characters only one edit distance (addition, deletion, or substitution) apart from the orthographic form of the correct translation (e.g., "pengiun" instead of "penguin"), as long as the response did not correspond to a distinct English word. A response was coded as *False friend* if the participant provided a phonologically similar incorrect translation. Responses not meeting the criteria for previous categories were labelled as *Wrong* or *Other* (see Data analysis section for more details). Both *Correct* and *Typo* responses were considered as correct, while *Wrong* and *False friend* responses were considered as incorrect. *Other* responses were excluded from data analysis. Trials in which participants took longer than 10 seconds to respond were also excluded. Participants contributed a total of `r format(nrow(exp_responses), big.mark = ",")` valid trials (`r format(nrow(exp_responses[exp_responses$group %in% c("cat-ENG", "cat-SPA"),]), big.mark = ",")` in Catalan, `r format(nrow(exp_responses[exp_responses$group %in% c("spa-ENG"),]), big.mark = ",")` in Spanish). The task took approximately 15 minutes to be completed.

@@

::: {.content-visible when-format="pdf"}
$$
\begin{aligned}
&\textbf{Likelihood}  \\
y_{i} \sim & \text{Bernoulli}(p_{i}) \\ \\
&\textbf{Parameters}  \\
\text{Logit}(p_{i}) = &  \beta_{0[p,w]} + \beta_{1[p]} \text{Frequency}_{i} + \beta_{2[p]} \text{PTHN}_i + \beta_{3[p]} \text{Similarity}_i + \beta_{4[p]} (\text{PTHN}_i \times \text{Similarity}_i) \\
\beta_{0-6[p,w]} \sim & \mathcal{N}(\mu_{\beta_{j}}, \sigma_{\beta_{j}}) \text{, for participant } p \text{ in 1, ..., } P \text{ and  word } w \text{ in 1, ..., } W \\
\beta_{1-6[p]} \sim &  \mathcal{N}(\mu_{\beta_{j}}, \sigma_{\beta_{j}}) \text{, for participant } p \text{ in 1, ..., } P \\ \\
&\textbf{Prior}  \\
\mu_{\beta_{p,w}}  \sim &  \mathcal{N}(0, 0.1) \\
\sigma_{\beta_{p}},  \sigma_{\beta_{w}} \sim & \text{HalfCauchy}(0, 0.1) \\
\rho_{p}, \rho_{w} \sim & \text{LKJCorr}(8) \\
\end{aligned}
$$ {#eq-1}

To test the practical relevance of each predictor we followed @kruschke2018bayesian. We first specified a region of practical equivalence (ROPE) around zero ([-0.1, +0.1], in the logit scale). This area indicates the values of the regression coefficients that we considered as equivalent to zero. We then computed the 95% posterior credible intervals (CrI) of the regression coefficients of interest. Finally, we calculated the proportion of the 95% CrI that fell inside the ROPE. This proportion indicates the probability that the true value of the coefficient is equivalent to zero. All analyses were performed in R environment [@rcoreteam2013language]. We used the tidyverse family of R packages [@wickham2019welcome] to process data and to generate figures. We used the `brms` R package [@burkner2017brms] using the `cmdstanr` back-end to the Stan probabilistic language [@carpenter2017stan] to estimate and compare the models (see Appendix 1 for mode details on the models).

@@

We collected data for a total of `r format(nrow(dataset$cat_eng) + nrow(dataset$spa_eng), big.mark = ",")` trials completed by `r n_participants$cat_eng + n_participants$spa_eng` unique participants. Of those trials, `r format(nrow(dataset$cat_eng), big.mark = ",")` were provided by `r n_participants$cat_eng` unique participants who listened to Catalan words, and `r format(nrow(dataset$spa_eng), big.mark = ",")` trials were provided by `r n_participants$spa_eng` unique participants who listened to Spanish words. We excluded trials in which participants did not enter any text (*n* = `r r_validity$cat_eng["blank"]`), in which a response in a language other than English was provided (e.g., "`agua`", *n* = `r r_validity$cat_eng["language"]`), in which participants did not provide a whole word (e.g., "`f`", *n* = `r r_validity$cat_eng["incomplete"]`), and in which participants added comments to the experimenter (e.g., "`unsure`", *n* = `r r_validity$cat_eng["other"]`). In addition, we excluded data from participants that self-rated their oral and/or written skills in Catalan and Spanish, or any other second language as four or higher in a five-point scale (*n* = `r excluded_lang`), were diagnosed with a language (*n* = `r excluded_impairment`), or did not contribute more than 80% of valid trials (*n* = `r excluded_ntrials`).

@@

@tbl-dataset shows a summary of participants' accuracy across Experiments 1, 2, and 3. MCMC chains in the model showed strong evidence of convergence ($\hat{R}<1.01$) (see Appendix 2 for more detailed model diagnostics). Participants translating Catalan words and participants translating Spanish words performed equivalently, as indicated by the regression coefficient of *Group* ($\beta$ = `r round(c1$b_group1$.median, 3)`, 95% CrI = [`r round(c1$b_group1$.lower, 3)`, `r round(c1$b_group1$.upper, 3)`], *p*(ROPE) = `r round(c1$b_group1$.rope, 3)`). Overall, participants responded less accurately to words with more CLPNs than to words with fewer CLPNs, regardless of the amount of phonological similarity between the presented word and its translation. This is indicated by the size of the regression coefficient of the two-way interaction between *Similarity* and *CLPN* ($\beta$ = `r round(c1$b_neigh_n_h_lv$.median, 3)`, 95% CrI = [`r round(c1$b_neigh_n_h_lv$.lower, 3)`, `r round(c1$b_neigh_n_h_lv$.upper, 3)`], *p*(ROPE) = `r round(c1$b_neigh_n_h_lv$.rope, 3)`). As anticipated, participants' performance benefited from an increase in *Similarity* ($\beta$ = `r round(c1$b_lv$.median, 3)`, 95% CrI = [`r round(c1$b_lv$.lower, 3)`, `r round(c1$b_lv$.upper, 3)`], *p*(ROPE) = `r round(c1$b_lv$.rope, 3)`), while the number of *CLND* had the opposite effect ($\beta$ = `r round(c1$b_neigh_n_h$.median, 3)`, 95% CrI = [`r round(c1$b_neigh_n_h$.lower, 3)`, `r round(c1$b_neigh_n_h$.upper, 3)`], *p*(ROPE) = `r round(c1$b_neigh_n_h$.rope, 3)`). @fig-epreds-1 illustrates the posterior of the average predictions of the model for words with different values of *Similarity* and *CLPN*.

@@

We collected data for a total of `r format(nrow(dataset$cat_spa) + nrow(dataset$cat_spa), big.mark = ",")` trials completed by `r n_participants$cat_spa` unique participants. We excluded trials in which participants did not enter any text (*n* = `r r_validity$cat_spa["blank"]`), in which a response in a language other than Spanish was provided (*n* = `r r_validity$cat_spa["language"]`), in which participants did not provide a whole word (*n* = `r r_validity$cat_spa["incomplete"]`), and in which participants added comments to the experimenter (*n* = `r r_validity$cat_spa["other"]`). In addition, we excluded data from participants that self-rated their oral and/or written skills in Catalan and Spanish, or any other second language as four or higher in a five-point scale (*n* = `r excluded_lang`), were diagnosed with a developmental language disorder (*n* = `r excluded_impairment`), or did not contribute more than 80% of valid trials (*n* = `r excluded_ntrials`). After applying trial-level and participant-level inclusion criteria, the resulting dataset included `r format(nrow(dataset_clean$cat_spa), big.mark = ",")` trials provided by `r n_distinct(dataset_clean$cat_spa$participant_id) + n_distinct(dataset_clean$cat_spa$participant_id)` unique participants. Of those trials, `r format(nrow(dataset_clean$cat_spa), big.mark = ",")` were provided by `r n_distinct(dataset_clean$cat_spa$participant_id)` unique participants who listened to Catalan words, and `r format(nrow(dataset_clean$cat_spa), big.mark = ",")` trials were provided by `r n_distinct(dataset_clean$cat_spa$participant_id)` unique participants who listened to Spanish words. Responses given by participants were `r round(lengths$len_mean, 2)` characters long on average (*Median* = `r round(lengths$len_median, 2)`, *SD* = `r round(lengths$len_sd, 2)`, Range = `r round(lengths$len_min, 2)`-`r round(lengths$len_max, 2)`).

```{r posterior-2}
c2 <- gather_draws(exp_2_m0, `b_.*`, `sd_.*`, regex = TRUE) |>
  summarise(across(.value, lst(
    .median = median,
    .lower = \(x) hdi(x)[1],
    .upper = \(x) hdi(x)[2],
    .rope = \(x) mean(between(x, -0.1, 0.1))
  ),
  .names = "{fn}"
  ))
c2 <- split(c2, c2$.variable)
names(c2) <- janitor::make_clean_names(names(c2))
```

@@

```{r tbl-surprises}
#| label: tbl-surprises
#| tbl-cap: "List of items with unexpectedly high accuracy: the Levenshtein similarityscore betwen the presented word (in Catalan or Spanish) and their correct Enlgish translation is zero, but participants, who are reportedly unfamiliar with the presented language, were on average >10% likely to guess the correct translation."
tbl_data <- list(
  "Experiment 1" = dataset_1,
  "Experiment 2" = dataset_2
) |>
  bind_rows(.id = "exp") |>
  mutate(translation = paste0())
  summarise(
    n = n(),
    correct = sum(correct),
    .by = c(exp, group, lv, translation)
  ) |>
  mutate(
    accuracy = prop_adj(correct, n),
    accuracy_se = prop_adj_se(correct, n)
  ) |>
  dplyr::filter(lv == 0, accuracy >= 0.1) |>
  select(exp, translation, group, accuracy, accuracy_se) |>
  arrange(exp, group, desc(accuracy))

tbl_data |>
  select(translation, accuracy, accuracy_se) |> 
  mutate(accuracy = round(accuracy * 100, 2),
         accuracy_se = round(accuracy_se * 100, 2)) |> 
  set_names(c("Translation", "Accuracy (%)", "SE")) |> 
  tt() |> 
  group_tt(i = list(
    "Experiment 1 (cat-ENG)" = 5,
    "Experiment 2 (spa-ENG)" = 21,
    "Experiment 3 (cat-SPA)" = 26))
```

@@

@@

```{r participants-excluded-3}
#| label: participants-excluded-3
excluded_lang <- participants_3 |>
  filter(l2_writ_prod > 3 |
    l2_oral_comp > 3 |
    cat_oral_comp > 3 |
    cat_writ_prod > 3 |
    spa_oral_comp > 3 |
    spa_writ_prod > 3) |>
  nrow()

@@

```{r dataset-3}
#| label: dataset-3
dataset <- split(quest_responses, quest_responses$group)
names(dataset) <- janitor::make_clean_names(names(dataset))
r_validity <- map(
  dataset,
  \(x) janitor::clean_names(table(x$response_type))
)
n_participants <- map(dataset, \(x) n_distinct(x$participant_id))

dataset <- quest_responses |>
  filter(group != "cat-SPA") |>
  group_split(group) |>
  set_names(c("spa-ENG", "cat-ENG")) |>
  janitor::clean_names()

@@

We collected data for a total of `r format(nrow(dataset$cat_eng) + nrow(dataset$spa_eng), big.mark = ",")` trials completed by `r n_participants$cat_eng + n_participants$spa_eng` unique participants. Of those trials, `r format(nrow(dataset$cat_eng), big.mark = ",")` were provided by `r n_participants$cat_eng` unique participants who listened to Catalan words, and `r format(nrow(dataset$spa_eng), big.mark = ",")` trials were provided by `r n_participants$spa_eng` unique participants who listened to Spanish words. We excluded trials in which participants did not enter any text (*n* = 0), in which a response in a language other than English was provided (*n* = `r r_validity$cat_eng["language"]`), in which participants did not provide a whole word (*n* = 0), and in which participants added comments to the experimenter (*n* = 0). In addition, we excluded data from participants that self-rated their oral and/or written skills in Catalan and Spanish, or any other second language as four or higher in a five-point scale (*n* = `r excluded_lang`), were diagnosed with a language (*n* = `r excluded_impairment`), or did not contribute more than 80% of valid trials (*n* = `r excluded_ntrials`).

@@

```{r lengths-2}
lengths <- dataset_3 |>
  filter(!knowledge) |> 
  mutate(len = map_int(strsplit(response, ""), length)) |>
  summarise(
    across(
      len,
      lst(mean, median, sd, min, max)
    ),
    .by = c(group)
  )
lengths <- split(lengths, lengths$group)

@@

```{r fig-knowledge}
#| label: fig-knowledge
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "Catalan and Spanish prior word knowledge, as reported by English native participants in Experiment 3. The X-axis indicates the proportion of participants that reported prior knowledge with the word. The Y-axis indicates participants' accuracy (the proportion of correct responses by participants). Each colour indicates whether the data point corresponds to a word that participants answered to with expected accuracy (red) or with unexpectedly high accuracy (blue), as defined by (no phonological similarity with the correct translation, by accuracy higher than 10%)."
p <- dataset_3 |>#| 
  mutate(surprise = translation %in% tbl_data$translation) |>
  summarise(
    knowledge = mean(knowledge),
    correct = sum(correct),
    n = n(),
    .by = c(translation, group, surprise)
  ) |>
  mutate(accuracy = prop_adj(correct, n),
  surprise = if_else(
    surprise,
    "Unexpected high accuracy\n(no phonological similarity)",
    "Expected accuracy"
  )) |> 
  ggplot(aes(knowledge, accuracy, colour = surprise, fill = surprise, shape = surprise)) +
  facet_wrap(~group) +
  geom_point(stroke = 1, size = 2) +
  #geom_rug(outside = TRUE, sides = "tr") +
  labs(
    x = "Reported knowledge",
    y = "Accuracy"
  ) +
  scale_x_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_shape_manual(values = c(1, 2)) +
  scale_colour_manual(values = c(
    RColorBrewer::brewer.pal(3, "Reds")[3], 
    RColorBrewer::brewer.pal(3, "Blues")[3]
  )) +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
  )
```

```{r draws-3}
c3 <- gather_draws(exp_3_m1, `b_.*`, `sd_.*`, regex = TRUE) |>
  summarise(across(.value, lst(
    .median = median,
    .lower = \(x) hdi(x)[1],
    .upper = \(x) hdi(x)[2],
    .rope = \(x) mean(between(x, -0.1, 0.1))
  ),
  .names = "{fn}"
  ))
c3 <- split(c3, c3$.variable)
names(c3) <- janitor::make_clean_names(names(c3))
```
